---
title: "segRDA vignette "
author:
  Danilo Cândido Vieira;
  Marco Colossi Brustolin;
  Fabio Cop;
  Gustavo Fonseca.
date: June 22, 2018
output:
  html_vignette:: rmarkdown::html_vignette
  html_document:
    theme: readable
    highlight: textmate
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{segRDA}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
```{r include = FALSE}
library(segRDA)
```

`segRDA` is designed to model non-continuous linear responses of ecological communities to environmental data. This tutorial assumes familiarity both with R and with community ordination. The package `segRDA` is available on CRAN or the latest development version can be installed from GitHub using the package `devtools` (\link{https://github.com/DaniloCVieira/segRDA}).


```{r Install, eval=FALSE}
# to install the package via CRAN
install.packages("segRDA")

# to install the package via GitHub
devtools::install_github("DaniloCVieira/segRDA")

```

The modeling process with `segRDA` is straightforward through three steps: **(1) data ordering** (function `OrdData`), **(2) split-moving-window analysis** (function `SMW`) and **(3) piecewise redundancy analysis** (function `pwRDA`). `segRDA` provides example data sets to facilitate reproducible examples. Through the following code we will use the dataset `sim3` from the package, which is a list composed of two matrices: `envi` and `comm`.


```{r }
data(sim3) ##Simulated data
x<-sim3$envi ## matrix of explanatory variables
y<-sim3$comm ## matrix of response variables

```

# Data ordering 
Both the SMW and pwRDA analyses depend on ordered datasets. The ordering can be user-defined or generated using `OrdData` function. The latter is a wrapper for (1) performing a redundancy analysis using the `rda` function from `vegan` package and (2) ordering both response and explanatory matrices using one of the axes of the RDA model. Defaults to use the first RDA axis. `rda` arguments can be passed directly to the `OrdData` function.


```{r Data ordering, fig.height=4, fig.width=8,fig.show='hold'}
sim1o<-OrdData(x=x,y=y, axis=1)
xo<-sim1o$xo ## ordered explanatory matrix
yo<-sim1o$yo ## ordered response matrix
```

The resulting ordered community can be represented by a species-site interaction matrix using the function `image`, which is available from the stats package of R.
```{r, fig.height=4, fig.width=8,fig.show='hold', tidy=TRUE, tidy.opts=list(width.cutoff=60)}

par(mfrow=c(1,2), mgp=c(1,1,0))
image(y, main="Original community data", col=topo.colors(100), axes=F, xlab="Sites", ylab="Species abundance")
image(yo, main="Ordered comunity data", col=topo.colors(100), axes=F, xlab="Sites")
```

# Split-moving window analysis

## A brief overview of the SMW analysis.

The application of SMW applied to a species matrix allows consists of (1) placing a window of even-numbered size at the beginning of the data series, (2) splitting the window into two equal halves, (3) calculating the community centroids within each half, (4) computing a dissimilarity metric between the two halves, (5) shifting window one position along the series, and (6) repeating the procedure till the end of the data series  (Cornelius & Reynolds 1991). The significance of the dissimilarity values within each window is tested through multi-response permutations and includes two types of procedures: the **“random-plot”** and **“random-shift”**. The first randomizes sites along the series while preserving species composition and abundance structure of the sites. The second randomizes patterns of each species relative to each other while preserving the spatial structure of the species abundance. Therefore, choosing between random plot and random shift relies on considering sites or species as a fixed aspect of the null distribution. Both procedures include computing an expected mean dissimilarity (DS) and standard deviation (SD) for each window midpoint. Confidence limits have been suggested as one or two SD above DS or estimated from one-tailed 95% confidence intervals (Erdos et al. 2014).

The choice of window size affects the results of the SMW analysis.  A *pooled* profile by averaging together dissimilarities from different window sizes reduces the scale-effect. In this case, dissimilarity profiles of the mean Z-score (standardized values of dissimilarity) are used to detect the community breakpoints. 

In plot randomization, Z-scores increase at each window midpoint together with an increase in window width, but the general shape of the dissimilarity profiles remains unchanged. In contrast, the random-shift method minimizes the scale dependence of Z-scores and makes a clear distinction between non-significant and significant discontinuities (Körmöczi et al. 2016).

## The function SMW
The package `segRDA` implements the SMW analysis through the function `SMW` which requires the following arguments:

* `yo`: the ordered community;
* `ws`: either a single window size or a vector of several windows sizes;
* `dist`: the distance metric (all distances metrics available in `vegan::vegdist`);
* `rand`: the randomization type (`"shift"` or `"plot"`);
* `n.rand`: the number of randomizations;

The default settings are `dist="bray"`, `rand="shift"` and `n.rand=99`. The running time of the analysis will depend on the number of window sizes (`length(ws)`) multiplied by the number of randomizations `n.rand`. In the following examples, we settled a low number of randomizations to speed up the analyses with the remaining arguments left at their default values.

```{r a_ws, results='hide',collapse = TRUE}
ws20<-SMW(yo=yo,ws=20, n.rand=10)
pool<-SMW(y=yo,ws=c(10,20,30,40), n.rand=10)
```

 `SMW` returns invisibly an `smw` object, which is a two-level list object describing the SMW results for each window `w` analyzed.  Each of the `w` slots is a list containing the following results:
 
* `..$dp`: The dissimilarity profile table (DP): a data frame giving the positions, labels, values of dissimilarity and z-scores for each sample;
*	`..$rdp `: data frame containing the randomized DP;
*	`..$md`: mean dissimilarity of the randomized DP;
*	`..$sd`: standard deviation for each sample position;
*	`..$oem`: overall expected mean dissimilarity;
*	`..$osd`: average standard deviation for the dissimilarities;
*	`..$params`: list with input arguments.


```{r collapse = T}
class(ws20)
length(ws20)
names(ws20)

class(pool)
length(pool)
names(pool)
```

The `smw` object have its own generic `print` method, `print.smw`, which is a shortcut for displaying the DP. 
```{r collapse = T}
class(ws20)
head(print(ws20))
```

## Methods for extracting  and plotting the SMW results
### `extract`
The function `extract` allows the user to access the results cointained in a `smw` object. The argument `index` specifies which of the results described above will be extracted. By default, `extract` returns an object of class `dp`: a DP table containing significant discontinuities and suggested breakpoints. Extracting a `dp` object implements the following auxiliary arguments:

* `sig`: defines the statistical test used to detect significant discontinuities. The statistical tests available are `"z"`,`"sd"`,`"sd2"`, and `"tail1"`;
* `z`: sets the critical value valor for z-scores when `sig="z"`;
* `BPs`: defines if the breakpoints should be chosen as those site positions corresponding to the maximum dissimilarity in a sequence of significant dissimilarity values (`BPs="max"`) or as those site positions corresponding to the median position of the sequence (`BPs="median"`). If `NULL` the breakpoints are not computed;
* `seq.sig`: specifies the length of a sequence of significant dissimilarity values that will be considered in defining the community breakpoints.

The default settings are `index="dp"`, `sig="z"`, `z="1.85"`, `BPs="max"` and `seq.sig=3`. The returned `dp` object has its own generic `print` method, `print.dp`:
```{r collapse = T}
ws20_dp<-extract(ws20) 
head(print(ws20_dp))
```

When the `smw` object handles multiple window sizes (hereafter referred to as *pooled* `smw` objects), the extracted DP table will contain the z-scores averaged over the set of window sizes.
```{r collapse = T}
pool_dp<-extract(pool) 
head(print(pool_dp))
```

A specific DP can be extracted from *pooled* objects by specifying a target window size (argument `w`).
```{r collapse = T}
ws10_dp<-extract(pool, w=10) 
```

The length of consecutive, significant dissimilarity values (argument `seq.sig`) is used as criteria for defining breakpoints. The function will display a warning if `seq.sig` exceeds the maximum threshold:

```{r collapse = T}
ws20_dp<-extract(ws20, sig="tail1", seq.sig=20)
```


The function `bp` is an auxiliary tool for displaying the breakpoints positions in objects of class `dp`  (if present): 
```{r collapse = T}
bp(ws10_dp)
bp(pool_dp)
```

Note that all other results contained in `smw` objects (i.e. `rdp`, `md`, `sd`, `oem`, `osd` and `params` ) comprise results of single window sizes. Thus, the argument `w` must be specified to access them from *pooled* objects. 

```{r collapse = T}
extract(pool, w=10, index="osd")
```

### `plot`
The `smw` object has a generic `plot` method, `plot.smw`, which is a convenience wrapper for accessing and viewing the results cointained in a `smw` object. Auxiliary arguments from `extract` (i.e.  `sig`, `z`, `BPs` and `seq.sig`) can be passed to `plot.smw`. The graphical output is the location of the window midpoints  vs. dissimilarity values. By default, significant dissimilarity values are given in "red" and the breakpoints in "blue". The function invisibly returns a default DP table (object class `dp`).

The background area of the graph can be coloured according to the breakpoints location (argument `bg`). These colors are  controled by the arguments `bg` and `bg_alpha`. The "bg" color generated for each sample can be acessed through the command `attr(dp,"params")$bg` (where 'dp' is an object of class `dp`). As we shall see later, this vector may be useful in the simultaneous visualization of the results the `SMW` and `pwRDA`. For the complete list of graphical parameters, please go to `help(plot.smw)`.

```{r collapse=T,fig.height=4, fig.width=8,fig.show='hold'}
par(mfrow=c(1,2))
plot(pool,w=20, main="DP from a single window (10)", cex.main=.8) 
plot(pool, main="DP from pooled windows (10, 20, 30 and 40)",  bg=c("rainbow"),cex.main=.8)

```

For *pooled* `smw` objects, a second type of plot is available: the window size effect. When the `plot` command is applied on a *pooled* `smw` object and the argument `w.effect` is `TRUE`, the `plot` method drawns together the profiles obtained from the different window sizes. 

```{r collapse=TRUE,fig.height=4, fig.width=4,fig.show='hold'}
plot(pool, w.effect = TRUE, main="Window size effect")
```

# Piecewise redundancy analysis (pwRDA) 
Once the breakpoints have been defined and their significance tested, the pwRDA can be implemented. It is done with the function `pwRDA`. The function has three arguments: `x` explanatory variables, `y` response variables and `BPs` the positions of the breakpoints. The function displays a message with the main result of the analysis:

```{r include=F}
pw.sim<-pwRDA(x.ord=xo,y.ord=yo, BPs=bp(pool_dp))
```
```{r eval=F}
pw.sim<-pwRDA(x.ord=xo,y.ord=yo, BPs=bp(pool_dp))
```
The function returns an invisible list with following descriptors:

* `..$summ`: a vector containing the summarized statistics.
* `..$rda.0` and `..$rda.pw`:  the respectives `cca` objects from  the full and "pw"" RDA models. These objects can be used in all functions that applies to `cca.object`. More details for accessing and plotting a `cca.object` are in `help(cca.object){vegan}` and `help(plot.cca){vegan}`, respectivaly. 

Finally, the `bg` vector obtained from the command `plot.smw` can be used for a simultaneous view of the results of the `SMW` and `pwRDA` analyses.

```{r fig.height=4, fig.width=12, collapse=TRUE,fig.show='hold'}
head(pw.sim$summ)
par(mfrow=c(1,3), cex=1.1)
# plotting the full rda model:
plot(pw.sim$rda.0, main="full RDA model", las=1)
# plotting the DP profile and keeping the sample colors:
dp<-plot(pool, main="DP from pooled windows (10, 20, 30 and 40)",  bg='rainbow',cex.main=.8)
bg.cols<-attr(dp,"params")$bg
# plotting the pwRDA colored according to the breakpoints:
plot(pw.sim$rda.pw,type="n", scaling=3, main="pwRDA model")
points(pw.sim$rda.pw, pch=16, col=bg.cols, cex=1.2)
text(pw.sim$rda.pw,  display="bp",pch=16,col="steelblue4",lwd=2)
```

# Literature
1. Burrough, P.A. (1986). Principles of Geographical Information Systems for Land Resources Assessment. Journal of Quaternary Sciencee, 193.

2. Cornelius, J.M. & Reynolds, J.F. (1991). On Determining the Statistical Significance of Discontinuities with Ordered Ecological Data. Ecology, 72, 2057–2070.

3. Erdos, L., Bátori, Z., Tölgyesi, C.S. & Körmöczi, L. (2014). The moving split window (MSW) analysis in vegetation science - An overview. Applied Ecology and Environmental Research, 12, 787–805.

4. Körmöczi, L., Bátori, Z., Erdős, L., Tölgyesi, C., Zalatnai, M. & Varró, C. (2016). The role of randomization tests in vegetation boundary detection with moving split-window analysis. Journal of Vegetation Science, 27, 1288–1296.

5. Oksanen, J., Blanchet, F.G., Friendly, M., Kindt, R., Legendre, P., McGlinn, D., Minchin, P.R., O’Hara, R.B., Simpson, G.L., Solymos, P., Stevens, M.H.H., Szoecs, E. & Wagner, H. (2017). vegan: Community Ecology Package. 
